---
title: "Trend of School Bus Delay Over Time"
subtitle: "Bike use and bike and microvehicles crashes over time"
date: 
output: 
  html_document:
    code_folding: hide
    always_allow_html: true
---

# Trends of School Bus Delay Over School years

## School Bus Delay Frequency and Time over School Years in Manhattan

On this page we will give you a bird-eye-view of school bus delay in Manhattan from 2018 Fall to 2021 Fall. We aims to explore the trends of the school bus delay occurrence and delay time over the school years. The time period we chose includes both pre and post covid data which will be interesting to look at ! Now lets dive right into it!

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "right"))

Sys.setlocale("LC_TIME", "English")
```

# Daily Delay Counts : from 9/15/2018 - 11/18/2021

```{r}
#read data in
df = read_csv("data/clean_data.csv")

#plot data: drop NA rows
plot_df = df %>% 
  select(school_year, year, how_long_delayed,occur_date, day, month,busbreakdown_id) %>% 
  na.omit()

#plot daily delay counts

plot_df %>% 
  group_by(occur_date) %>% 
  summarise(delay_count = n(),school_year) %>% 
  ggplot(aes(x = occur_date, y = delay_count)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  facet_grid(~school_year,scales = "free") + 
  labs(
    title = "Daily School Bus Delay Counts in Manhattan",
    x = "Date",
    y = "Counts"
  ) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1))
  


```


Lets Look at The Distribution of Delay Times in Each School Year

```{r}

delay_time_dist = 
  plot_df %>% 
  ggplot(aes(x = how_long_delayed,fill = how_long_delayed))+
    geom_bar(position = "dodge")+
  facet_grid(~school_year,scales = "free") + 
  labs(
    title = "Distribution of School Bus Delay Time in Manhattan",
    x = "Delay Time",
    y = "Counts"
  )+
  theme(axis.title.x = element_blank(),axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 0.5)) + 
  guides(fill=guide_legend(title="Delay Time"))

ggplotly(delay_time_dist)
```

lets now overlap the school bus delay counts within each school year 

```{r}
#overlaping each year's daily delay counts

overlaping_plot = plot_df %>% 
 group_by(occur_date) %>% 
  mutate(delay_count = n(),
         day_of_month = lubridate::mday(occur_date),
         common_day = lubridate::mdy(paste(month, day_of_month,"2021")))%>%
  ggplot(aes(x = common_day, y = delay_count, group = school_year, color = school_year)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) + 
  labs(
    title = "School Bus Delay in Manhattan",
    x = "Month",
    y = "Delay Counts"
    ) +
  scale_x_date(date_breaks = "1 month", labels = function(x) format(x, "%b"))+ guides(fill=guide_legend(title="School Year"))

ggplotly(overlaping_plot)
```



Plot4:
```{r}
df_date_weather_na = 
  df %>%
  drop_na(how_long_delayed) %>%
  mutate(
    month = forcats::fct_relevel(as.factor(month), "September", "October", "November", "December", "January", "February", "March", "April", "May", "June"),
    day = forcats::fct_relevel(as.factor(day), "Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
  ) %>%
  filter(!(school_year == "2021-2022"))
df_date_weather_na

df_date_weather_plot = 
  df_date_weather_na %>%
  group_by(month, how_long_delayed, school_year) %>%
  summarize(n_obs = n()) %>%
  mutate(how_long_delayed = forcats::fct_relevel(as.factor(how_long_delayed), "0-15 Min", "16-30 Min", "31-45 Min", "46-60 Min", "61-90 Min")
         )

p1 = ggplot(df_date_weather_plot, aes(x = month, y = n_obs)) + 
  geom_col(alpha = 0.6) + 
  facet_grid(~how_long_delayed) +
  geom_line(aes(group = school_year, color = school_year)) + 
  labs(
    x = "month",
    y = "counts",
    title = "the monthly distribution for different length of time"
  ) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
ggplotly(p1)
```



Plot5:
```{r}
df_date_weather_plot2 =
  df_date_weather_na %>%
  group_by(day, school_year, how_long_delayed) %>%
  summarize(n_obs = n())

p2 = ggplot(df_date_weather_plot2, aes(x = day, y = n_obs)) + 
  geom_col(alpha = 0.6) + 
  facet_grid(~school_year) +
  labs(
    x = "weekday",
    y = "counts",
    title = "the daily distribution for different length of time"
  ) + 
  geom_line(aes(group = how_long_delayed, color = how_long_delayed)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
ggplotly(p2)
```

